---
format: gfm
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
devtools::load_all(quiet = TRUE)
```

# q2r

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

`q2r` is an exploratory R package that wraps the [`pampa`](https://github.com/quarto-dev/q2)
Rust crate to expose Quarto's QMD parser to R. It parses QMD text or files and
returns either the tree-sitter AST or the Pandoc AST, together with any parse
diagnostics.

## Installation

`q2r` is not on CRAN. You can install the development version from GitHub with:

``` r
# install.packages("pak")
pak::pak("rundel/q2r")
```

Building from source requires a working Rust toolchain with `rustc >= 1.85`
(edition 2024). [`rustup`](https://rustup.rs/) is the most reliable way to get
this. The first build fetches the upstream `q2` workspace (~40 crates) via
cargo; subsequent builds use the cargo cache.

## Usage

The single entry point is `parse_qmd()`. With `ast = "pd"` (the default) it
returns a `pandoc` S7 object holding the Pandoc AST; with `ast = "ts"` it
returns a `ts_tree` S7 object holding the tree-sitter concrete syntax tree. It
accepts either a file path or a string of QMD text, and attaches any parse
diagnostics to the returned object's `@diagnostics` slot.

```{r}
#| eval: false
library(q2r)
```

```{r}
qmd = "---
title: Example
---

# Heading

Some *emphasized* text with a [link](https://example.com).
"
```

### Pandoc AST

```{r}
pd = parse_qmd(qmd)
pd
```

### Tree-sitter AST

```{r}
ts = parse_qmd(qmd, ast = "ts")
ts
```

### Rendering back to QMD

`to_qmd()` renders a parsed AST (a `pandoc` or `ts_tree` object) back to QMD
source. A `pandoc` object is written by pampa's QMD writer; a `ts_tree` is
recovered from its source bytes.

```{r}
cat(to_qmd(pd))
```

### Diagnostics

Parse errors and warnings are returned as structured `pampa_diagnostic` objects
attached to the parsed result. By default `parse_qmd()` signals error-kind
diagnostics as R errors and warning-kind diagnostics as R warnings; pass
`quiet = TRUE` to suppress signalling and inspect them directly.

```{r}
#| error: true
bad = parse_qmd(
  "# Heading {.cls bad}

See [my page](https://example.com/some path) for details.
"
)
```

## Working with the AST

Beyond `parse_qmd()` / `to_qmd()`, q2r ships a tidyselect-style vocabulary for
querying and rewriting either AST (`select_nodes()`, `select_descendants()`,
`map_nodes()`, `replace_nodes()`, `delete_nodes()`, `insert_before()`, ...), an
`ast_filter()` Lua-filter-style walker, document-level helpers (`ast_summary()`,
`select_section()`, `ast_toc()`, `split_sections()`), code-cell helpers
(`cell_options()`, `set_cell_options()`, `collect_code()`), table bridges
(`as_df()` / `as_table()`), file round-trip sugar (`read_qmd()`, `write_qmd()`,
`edit_qmd()`), and multi-document collections (`parse_qmd_dir()`). See the
`vignette("selecting-and-rewriting")` for a tour.

## Related

- [`quarto-dev/q2`](https://github.com/quarto-dev/q2) - upstream Rust workspace
  containing `pampa` and the tree-sitter-qmd grammar.
