---
format: gfm
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
devtools::load_all(quiet = TRUE)
```

# q2r

<!-- badges: start -->
<!-- badges: end -->

`q2r` is an exploratory R package that wraps the [`pampa`](https://github.com/quarto-dev/q2)
Rust crate to expose Quarto's QMD parser to R. It parses QMD text or files and
returns either the tree-sitter concrete syntax tree or the Pandoc AST, together
with any parse diagnostics.

## Installation

`q2r` is not on CRAN. You can install the development version from GitHub with:

``` r
# install.packages("pak")
pak::pak("rundel/q2r")
```

Building from source requires a working Rust toolchain with `rustc >= 1.85`
(edition 2024). [`rustup`](https://rustup.rs/) is the most reliable way to get
this. The first build fetches the upstream `q2` workspace (~40 crates) via
cargo; subsequent builds use the cargo cache.

## Usage

The two main entry points are `pampa_parse_pd()`, which returns a `pandoc` S7
object holding the Pandoc AST, and `pampa_parse_ts()`, which returns a `ts_tree`
S7 object holding the tree-sitter concrete syntax tree. Both accept either a
file path or a string of QMD text, and attach any parse diagnostics to the
returned object's `@diagnostics` slot.

```{r}
#| eval: false
library(q2r)
```

```{r}
qmd = "---
title: Example
---

# Heading

Some *emphasized* text with a [link](https://example.com).
"
```

### Pandoc AST

```{r}
pd = pampa_parse_pd(qmd)
pd
```

### Tree-sitter AST

```{r}
ts = pampa_parse_ts(qmd)
ts
```

### Rendering back to QMD

`pampa_to_qmd()` renders a parsed AST (or text / file path) back to QMD source
via pampa's writer.

```{r}
cat(pampa_to_qmd(pd))
```

### Diagnostics

Parse errors and warnings are returned as structured `pampa_diagnostic` objects
attached to the parsed result. By default `pampa_parse_*()` signals error-kind
diagnostics as R errors and warning-kind diagnostics as R warnings; pass
`quiet = TRUE` to suppress signalling and inspect them directly.

```{r}
bad = pampa_parse_pd("::: {.callout-note\nunterminated\n", quiet = TRUE)
bad@diagnostics
```

## Related

- [`quarto-dev/q2`](https://github.com/quarto-dev/q2) — upstream Rust workspace
  containing `pampa` and the tree-sitter-qmd grammar.
