---
title: "Selecting and rewriting AST nodes"
format: html
vignette: >
  %\VignetteIndexEntry{Selecting and rewriting AST nodes}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

q2r exposes two ASTs for a QMD document: the high-level [`pandoc`] S7
hierarchy and the [`ts_tree`] tree-sitter concrete syntax tree. The
`select_*` / `map_nodes` / `*_nodes` API lets you query and rewrite
either one with the same vocabulary.

```{r}
library(q2r)

src = "
# Intro

Some **bold** prose with a [link](https://example.com).

::: callout-note
## Watch out

A callout with its own *emphasised* heading.
:::

## Conclusion

Final words.
"

doc = pampa_parse_pd(src)
```

## Selecting

Predicates are unquoted R expressions evaluated against each candidate
node with a node-aware data mask. The mask exposes node slots
(`level`, `url`, `text`, `kind`, ...) and a small set of helpers
(`is()`, `has_class()`, `has_id()`, `has_attr()`, `is_leaf()`).

```{r}
select_nodes(doc, is(pandoc_header))
```

`is()` honours S7 inheritance, so `is(pandoc_block)` matches every
block kind:

```{r}
length(select_nodes(doc, is(pandoc_block)))
```

Bare slot access lets you filter on properties without writing
accessors yourself:

```{r}
select_nodes(doc, is(pandoc_header) & level == 2L)
```

`has_class()` reads `@attr@classes`:

```{r}
select_nodes(doc, is(pandoc_div) & has_class("callout-note"))
```

## Chaining

The selection verbs accept a list of nodes, so a pipe chain narrows
without nesting:

```{r}
doc |>
  select_nodes(is(pandoc_div) & has_class("callout-note")) |>
  select_descendants(is(pandoc_header))
```

`select_descendants()` excludes the root(s) of its input;
`select_children()` only inspects direct children.

## Rewriting

`map_nodes()` rewrites every match in place. `.f` may return a single
node (replace), a list of nodes (splice into the parent's child
list), or `NULL` (delete). The walker traverses bottom-up, matching
Pandoc Lua filters' default.

Promote every level-2 heading to level 1:

```{r}
out = map_nodes(
  doc,
  is(pandoc_header) & level == 2L,
  .f = \(h) pandoc_header(level = 1L, attr = h@attr, content = h@content)
)

purrr::map_int(select_nodes(out, is(pandoc_header)), \(h) h@level)
```

Strip bold by splicing each `pandoc_strong`'s content up into the
parent:

```{r}
out = map_nodes(doc, is(pandoc_strong), .f = ~ .x@content@content)
length(select_nodes(out, is(pandoc_strong)))
```

## Structural edits

`insert_before()` / `insert_after()` add siblings around each match;
`delete_nodes()` removes them; `replace_nodes()` swaps them out for a
constant value.

```{r}
banner = pandoc_paragraph(
  content = pandoc_inlines(list(pandoc_str(text = "[banner]")))
)

doc |>
  insert_before(is(pandoc_header) & level == 1L, .what = banner) |>
  select_children(is(pandoc_paragraph) | is(pandoc_header))
```

## The tree-sitter side

Every verb above works on a [`ts_tree`] too. The mask exposes the
ts-specific slots (`kind`, `is_named`, `field_name`, `text`):

```{r}
ts = pampa_parse_ts(src)

select_nodes(ts, kind == "atx_heading")
```

For full structural pattern matching, captures, and predicates
(`#eq?`, `#match?`, ...) there's a [`ts_query()`] escape hatch that
runs a tree-sitter `.scm` query directly:

```{r}
matches = ts_query(ts, "(atx_heading) @h")
length(matches)
matches[[1]]$h
```

## Round-trips

Rewriting produces a new AST of the same class as the input, so the
result feeds straight back into the renderer:

```{r}
out = doc |>
  map_nodes(
    is(pandoc_header) & level == 1L,
    .f = \(h) pandoc_header(level = 2L, attr = h@attr, content = h@content)
  )

cat(pampa_to_qmd(out))
```
