---
title: "Selecting and rewriting AST nodes"
format: html
vignette: >
  %\VignetteIndexEntry{Selecting and rewriting AST nodes}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

q2r exposes two ASTs for a QMD document: the high-level [`pandoc`] S7
hierarchy and the [`ts_tree`] tree-sitter concrete syntax tree. The
`select_*` / `map_nodes` / `*_nodes` API lets you query and rewrite
either one with the same vocabulary.

```{r}
library(q2r)

src = "
# Intro

Some **bold** prose with a [link](https://example.com).

::: callout-note
## Watch out

A callout with its own *emphasised* heading.
:::

## Conclusion

Final words.
"

doc = pampa_parse_pd(src)
```

## Selecting

Predicates are unquoted R expressions evaluated against each candidate
node with a node-aware data mask. The mask exposes node slots
(`level`, `url`, `text`, `kind`, ...) and a small set of helpers
(`is()`, `has_class()`, `has_id()`, `has_attr()`, `is_leaf()`).

```{r}
select_nodes(doc, is(pandoc_header))
```

`is()` honours S7 inheritance, so `is(pandoc_block)` matches every
block kind:

```{r}
length(select_nodes(doc, is(pandoc_block)))
```

Bare slot access lets you filter on properties without writing
accessors yourself:

```{r}
select_nodes(doc, is(pandoc_header) & level == 2L)
```

`has_class()` reads `@attr@classes`:

```{r}
select_nodes(doc, is(pandoc_div) & has_class("callout-note"))
```

## Chaining

The selection verbs accept a list of nodes, so a pipe chain narrows
without nesting:

```{r}
doc |>
  select_nodes(is(pandoc_div) & has_class("callout-note")) |>
  select_descendants(is(pandoc_header))
```

`select_descendants()` excludes the root(s) of its input;
`select_children()` only inspects direct children.

## Rewriting

`map_nodes()` rewrites every match in place. `.f` may return a single
node (replace), a list of nodes (splice into the parent's child
list), or `NULL` (delete). The walker traverses bottom-up, matching
Pandoc Lua filters' default.

Promote every level-2 heading to level 1:

```{r}
out = map_nodes(
  doc,
  is(pandoc_header) & level == 2L,
  .f = \(h) pandoc_header(level = 1L, attr = h@attr, content = h@content)
)

purrr::map_int(select_nodes(out, is(pandoc_header)), \(h) h@level)
```

Strip bold by splicing each `pandoc_strong`'s content up into the
parent:

```{r}
out = map_nodes(doc, is(pandoc_strong), .f = ~ .x@content@content)
length(select_nodes(out, is(pandoc_strong)))
```

## Structural edits

`insert_before()` / `insert_after()` add siblings around each match;
`delete_nodes()` removes them; `replace_nodes()` swaps them out for a
constant value.

```{r}
banner = pandoc_paragraph(
  content = pandoc_inlines(list(pandoc_str(text = "[banner]")))
)

doc |>
  insert_before(is(pandoc_header) & level == 1L, .what = banner) |>
  select_children(is(pandoc_paragraph) | is(pandoc_header))
```

## Type-keyed dispatch with `ast_filter()`

`map_nodes()` is most natural for single-type rewrites driven by a
predicate. When you want to rewrite *several* node types in a single
pass, `ast_filter()` accepts a table of handlers keyed by S7 class.
This is the q2r analogue of a Pandoc Lua filter
(`{ Header = function(el) ... end, Strong = ... }`).

```{r}
out = doc |> ast_filter(
  pandoc_strong = \(el) pandoc_small_caps(content = el@content),
  pandoc_header = \(el) {
    if (el@level == 1L) {
      pandoc_header(level = 2L, attr = el@attr, content = el@content)
    } else el
  }
)

cat(pampa_to_qmd(out))
```

Handlers honour the same return-value contract as `map_nodes()`: return
a node to replace, a list of nodes to splice, `NULL` to delete, or the
input to no-op.

Dispatch respects S7 inheritance, so an abstract-class handler acts as
a catch-all (first-match-wins, so put specific types before general
ones):

```{r}
visits = c()
invisible(doc |> ast_filter(
  pandoc_header = \(el) { visits <<- c(visits, "header"); el },
  pandoc_block  = \(el) { visits <<- c(visits, "block");  el }
))
table(visits)
```

## Top-down traversal and skipping subtrees

Pass `.order = "pre"` to visit parents before children. From a pre-order
handler, return `ast_skip(x)` to install `x` at that position without
descending into its children. This mirrors Pandoc Lua filters'
`traverse = 'topdown'` plus `return el, false`.

```{r}
# Strip ALL formatting inside callout divs, leaving the rest of the
# document alone.
out = doc |> ast_filter(
  .order = "pre",
  pandoc_div = \(d) {
    if (has_class(d, "callout-note")) {
      flattened = pandoc_div(
        attr    = d@attr,
        content = as_blocks(ast_text(d))
      )
      ast_skip(flattened)
    } else d
  }
)

cat(pampa_to_qmd(out))
```

Without `ast_skip()`, the walker would still descend into the rewritten
div's content; the skip is what makes "this whole subtree is now done"
explicit.

## List-level filters

Two special handler names dispatch on entire inline/block sequences
after their contents have been rewritten:

- `pandoc_inlines = \(xs) ...` runs once per `pandoc_inlines` wrapper.
- `pandoc_blocks = \(xs) ...` is the equivalent for `pandoc_blocks`.

These are essential for context-aware transforms that single-element
handlers cannot express (merging adjacent runs, dropping siblings based
on neighbours, etc.).

```{r}
# Drop any paragraph whose plain-text form contains "Final".
out = doc |> ast_filter(pandoc_blocks = \(xs) {
  keep = !purrr::map_lgl(xs@content, \(b) {
    S7::S7_inherits(b, pandoc_paragraph) &&
      grepl("Final", ast_text(b), fixed = TRUE)
  })
  pandoc_blocks(xs@content[keep])
})

select_children(out, is(pandoc_paragraph)) |>
  purrr::map_chr(ast_text)
```

List-level and element-level handlers compose in a single pass:

```{r}
out = doc |> ast_filter(
  pandoc_strong  = \(el) pandoc_small_caps(content = el@content),
  pandoc_inlines = \(xs) {
    # Append a marker whenever a paragraph ends up containing small caps.
    has_smallcaps = any(purrr::map_lgl(xs@content, S7::S7_inherits, pandoc_small_caps))
    if (has_smallcaps) {
      pandoc_inlines(c(xs@content, list(pandoc_str(text = "[!]"))))
    } else xs
  }
)

# the marker landed in the paragraph that had bold:
purrr::map_chr(select_children(out, is(pandoc_paragraph)), ast_text)
```

## Flattening to plain text with `ast_text()`

`ast_text()` recursively concatenates the textual content of any
subtree, dropping all formatting. It is the q2r analogue of
`pandoc.utils.stringify()` and is handy for matching on document
content without walking the AST yourself.

```{r}
ast_text(doc)
```

It works on any node, not just the root:

```{r}
doc |>
  select_first(is(pandoc_div) & has_class("callout-note")) |>
  ast_text()
```

A common pattern: pick out matches by predicate, then summarise:

```{r}
doc |>
  select_nodes(is(pandoc_header)) |>
  purrr::map_chr(ast_text)
```

## Editing attributes

The `@attr` slot carries id, classes, and key/value attributes for the
nodes that have one (headers, divs, code, links, spans, ...). q2r ships
immutable getters/setters that read like Lua's direct field access
(`el.classes`, `el.identifier`) but return a new node:

- `has_class(x, cls)` / `add_class(x, cls)` / `remove_class(x, cls)`
- `get_id(x)` / `set_id(x, id)`
- `get_attr(x, key)` / `set_attr(x, key, value)` / `remove_attr(x, key)`

```{r}
out = doc |> ast_filter(pandoc_header = \(h) {
  h |> add_class("section") |> set_attr("data-level", as.character(h@level))
})

out |> select_nodes(is(pandoc_header)) |> purrr::map(\(h) h@attr)
```

Nodes without an `@attr` slot degrade gracefully for predicates
(`has_class(pandoc_str(...), "x")` returns `FALSE`) but error on
setters with a clear message.

## Constructing content quickly

Pandoc constructors are strictly typed, so even a simple emphasised
phrase requires a few layers of wrapping. `as_inlines()` and
`as_blocks()` accept strings (whitespace is converted to spaces,
newlines to soft breaks), single nodes, lists, or existing wrappers,
and produce the canonical wrapper type. Use them inside an
`ast_filter()` handler to keep the body terse.

```{r}
# Append a "[Edited]" tag to every paragraph.
out = doc |> ast_filter(pandoc_paragraph = \(p) {
  pandoc_paragraph(content = as_inlines(c(
    p@content@content,
    list(pandoc_space(), pandoc_str(text = "[Edited]"))
  )))
})

select_children(out, is(pandoc_paragraph)) |>
  purrr::map_chr(ast_text)
```

`as_blocks()` makes the multi-paragraph case equally short:

```{r}
note = as_blocks(c(
  "First paragraph of an appended note.",
  "Second paragraph with more detail."
))

doc |>
  insert_after(is(pandoc_header) & level == 1L, .what = note@content) |>
  pampa_to_qmd() |>
  cat()
```

## The tree-sitter side

Every verb above works on a [`ts_tree`] too. The mask exposes the
ts-specific slots (`kind`, `is_named`, `field_name`, `text`):

```{r}
ts = pampa_parse_ts(src)

select_nodes(ts, kind == "atx_heading")
```

For full structural pattern matching, captures, and predicates
(`#eq?`, `#match?`, ...) there's a [`ts_query()`] escape hatch that
runs a tree-sitter `.scm` query directly:

```{r}
matches = ts_query(ts, "(atx_heading) @h")
length(matches)
matches[[1]]$h
```

## Round-trips

Rewriting produces a new AST of the same class as the input, so the
result feeds straight back into the renderer:

```{r}
out = doc |>
  map_nodes(
    is(pandoc_header) & level == 1L,
    .f = \(h) pandoc_header(level = 2L, attr = h@attr, content = h@content)
  )

cat(pampa_to_qmd(out))
```
